### 2021-11-17 text/outline/draft.tex
update text; set up simulations

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A new class of exponential family measurement error models, with application to single-cell CRISPR screens}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. In this work we study single-cell CRISPR screens from a statistical perspective. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses. Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models Poisson- or negative binomially-distributed gene expression counts. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on so-called high-multiplicity of infection (MOI) single-cell CRISPR screens. We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well. In this section we motivate high MOI single-cell screens, overview the experimental protocol, and present relevant analysis challenges.

The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current discussion). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes. We do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.

@@

\subsection{Zero-inflated model}
The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes. This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}. However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero. We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}. We derive efficient methods for estimation and inference in this model similar to those outlined above. However, given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.

>>>>

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A class of exponential family measurement error models, with application to single-cell CRISPR screens}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses. Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while regressing out technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens. We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well. In this section we motivate high MOI single-cell screens, overview the experimental protocol, and present relevant analysis challenges.

The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current discussion). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.

@@

\subsection{Zero-inflated model}
The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes. This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}. However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero. We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}. We derive efficient methods for estimation and inference in this model. Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.
